$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-21 paper/paper.tex
on typing and the example

@@

\begin{example}[The courier system]
\label{example-process}
\newcommand{\sender}{\textcolor{brown}{sender}}
\newcommand{\receiver}{\textcolor{purple}{receiver}}
\newcommand{\courier}{\textcolor{cyan}{courier}}

@@

In the linear \picalc{} each channel has an input and an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 (exhausted) or 1 (available).
We generalise over this notion of multiplicities by defining an algebra for them that is satisfied by linear, graded and shared types alike.
We then use pairs of multiplicities as usage annotations for channels.

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$, that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational resoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} models such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

The unit type $\unit$ serves as a proof of inhabitance for types.
The base type $\base{n}$ uses natural numbers as placeholders for types (the host language can then interpret the former into the latter --- e.g., $0$ as booleans, $1$ as natural numbers) so that we avoid having to deal with universe polymorphism.
The type $\channel{t}{x}$ of a channel determines what type ${t}$ of data and what usage annotations ${x}$ are sent over that channel.
This notation aligns with $[t]\ \mathbf{chan}_{(i^{y},o^{z})}$, where $y,z$ are the multiplicities of input and output, respectively \cite{K07}.

@@

Typing judgments for linear calculi typically carry a single context of usage annotations and thus the typing rules require frequent top-down proofs of context splitting --- e.g., the typing rule for parallel composition would look as follows:
\begin{mathpar}
  \inferrule
  {\opctx{\Gamma}{\Delta}{\Xi} \\ \Delta \; \type{\vdash} \; P \\ \Xi \; \type{\vdash} \; Q}
  {\Gamma \; \type{\vdash} \; \comp{P}{Q}}
\end{mathpar}
Input requires a similar context split; output requires two such context splits.
This is rather cumbersome: for input and output, the proof of context split carries \emph{exactly} the same information as the typing rule for variables; for parallel composition it means that the typing derivations of both subprocesses will have no context information whatsoever until the context split is determined.

Leftover typing solves this problem by inverting the flow of information so that \textbf{the typing derivations of the subprocesses determine the context split}.
As a result, usage information is threaded bottom-up through the typing derivation, and top-down context split proofs are no longer needed.
In addition to this, it allows a \emph{framing} (\autoref{thm:framing}) theorem to be stated, and for the \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) theorems to acquire a more general form.

@@

  The inaction process in rule $\PO$ does not change usage annotations.
  The scope restriction in rule $\new$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ of what is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that they are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $y \comma y$, which it must completely exhaust.

  The input process in rule $\recv{}{}$ requires a channel $chan_i$ at index $i$ with usage $\li$ available, such that data with type $t$ and usage $x$ can be sent over it.
  Note that the index $i$ is used in the syntax of the typed process.
  We use the leftovers $\Xi$ to type the continuation process, which is also provided with the received element --- of type $t$ and multiplicity $x$ at index $\constr{0}$.
  The received element $x$ must be completely exhausted by the continuation process.

  Similarly to input, the output process in rule $\send{}{}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the continuation process, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of the channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.

  Parallel composition in rule $\comp{}{}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ of the resources $P$ uses, we can use them to type $Q$ and save the user from having to provide an extrinsic proof of context split.
\end{nidefinition}

@@

We provide an example typing derivation for the processes defined in \autoref{example-process}.
For the sake of simplicity, we instantiate these processes with concrete variable references before typing them.

The $\func{recv}$ process receives data along the channel with index $0$, thus that variable needs to be of type $\channel{t}{u}$ for some $t$ and $u$.
After receiving twice, the process ends, and we should not be left with any unused multiplicities: $u$ needs to be $\lz$.
We will use graded types to keep track of the exact number of times communication happens.
Whatever the input multiplicity of the channel, we will consume $2$ of it and leave the remaining as leftovers.
Once the types are given, the typing derivation is completely syntax directed.

The $\func{send}$ process sends data along the channel with index $0$, thus that variable needs to be of type $\channel{t}{u}$ for some $t$ and $u$.
We instantiate $t$ (the type of data that we transmit) to $\channel{\unit}{\func{\omega}}$, as that is a value we can easily construct.
As per the type of the process $\func{recv}$, the transmitted multiplicities $u$ are $\lz$.
We will transmit once, thus using a single output multiplicity, and leave the rest as leftovers.

@@

Dually, the process doing the synchronisation expects input multiplicities for the channels shared with $\func{send}$ and output multiplicities for the channel shared with $\func{recv}$.
The typing derivation here is uniquely determined by the type and syntax of the process.

@@

We can now put these processes together.
Specifying the types of the channels is not required: Agda can infer them from the types of the processes.

>>>>

% Example
\newcommand{\sender}{\textcolor{brown}{sender}}
\newcommand{\receiver}{\textcolor{purple}{receiver}}
\newcommand{\courier}{\textcolor{cyan}{courier}}

@@

\begin{example}[The courier system]
\label{example-process}

@@

In the linear \picalc{} each channel has an input and an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 (exhausted) or 1 (available).
We generalise over this notion by defining an algebra for multiplicities that is satisfied by linear, graded and shared types alike.
We then use pairs of multiplicities as usage annotations for a channel's input and output capabilities.

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} models such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

The unit type $\unit$ serves as a proof of inhabitance for types.
The base type $\base{n}$ uses natural numbers as placeholders for types (the host language can then interpret the former into the latter --- e.g., $0$ as booleans, $1$ as natural numbers) so that we avoid having to deal with universe polymorphism.
The type $\channel{t}{x}$ of a channel determines what type ${t}$ of data and what usage annotations ${x}$ are sent over that channel.
This channel notation aligns with $[t]\ \mathbf{chan}_{(i^{y},o^{z})}$, where $y,z$ are the multiplicities of input and output, respectively \cite{K07}.

@@

Typing judgments for linear calculi typically carry a single context of usage annotations and thus the typing rules require frequent top-down proofs of context splitting --- e.g., the typing rule for parallel composition would look like:
\begin{mathpar}
  \inferrule
  {\opctx{\Gamma}{\Delta}{\Xi} \\ \Delta \; \type{\vdash} \; P \\ \Xi \; \type{\vdash} \; Q}
  {\Gamma \; \type{\vdash} \; \comp{P}{Q}}
\end{mathpar}
Input requires a similar context split; output requires two such context splits.
This is rather cumbersome: for input and output, the proof of context split carries \emph{exactly} the same information as the typing rule for variables; for parallel composition it means that the typing derivations of both subprocesses will have no context information whatsoever until the context split is determined.

Leftover typing solves this problem by inverting the flow of information so that \textbf{the typing derivations of the subprocesses determine the context split}.
As a result, usage information is threaded bottom-up through the typing derivation, and top-down context split proofs are no longer needed.
In addition, it allows \emph{framing} (\autoref{thm:framing}) to be stated, and \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) to acquire a more general form.

@@

  The inaction process in rule $\PO$ does not change usage annotations.
  The scope restriction in rule $\new$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ of what is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that they are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $y \comma y$, which it must completely exhaust.
%
  The input process in rule $\recv{}{}$ requires a channel $chan_i$ at index $i$ with usage $\li$ available, such that data with type $t$ and usage $x$ can be sent over it.
  Note that the index $i$ is used in the syntax of the typed process.
  We use the leftovers $\Xi$ to type the continuation process, which is also provided with the received element --- of type $t$ and multiplicity $x$ at index $\constr{0}$.
  The received element $x$ must be completely exhausted by the continuation process.
%
  Similarly to input, the output process in rule $\send{}{}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the continuation process, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of the channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.
%
  Parallel composition in rule $\comp{}{}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ of the resources $P$ uses, we can use them to type $Q$ and save the user from having to provide an extrinsic proof of context split.
\end{nidefinition}

@@

We provide the typing derivation for the courier system defined in \autoref{example-process}.
For the sake of simplicity, we instantiate these processes with concrete variable references before typing them.

The \receiver{} defined by the $\func{recv}$ process receives data along the channel with index $0$, thus that variable needs to be of type $\channel{t}{u}$ for some $t$ and $u$.
After receiving twice, the process ends, and we should not be left with any unused multiplicities: $u$ needs to be $\lz$.
We will use graded types to keep track of the exact number of times communication happens.
Whatever the input multiplicity of the channel, we will consume $2$ of it and leave the remaining as leftovers.
Once the types are given, the typing derivation is completely syntax directed.

The \sender{} defined by the $\func{send}$ process sends data along the channel with index $0$, thus that variable needs to be of type $\channel{t}{u}$ for some $t$ and $u$.
We instantiate $t$ (the type of data that we transmit) to $\channel{\unit}{\func{\omega}}$, as that is a value we can easily construct.
As per the type of the process $\func{recv}$, the transmitted multiplicities $u$ are $\lz$.
We will transmit once, thus using a single output multiplicity, and leave the rest as leftovers.

@@

Dually, the \courier{} defined by the $\func{carry}$ process expects input multiplicities for the channels shared with $\func{send}$ and output multiplicities for the channel shared with $\func{recv}$.
The typing derivation here is uniquely determined by the type and syntax of the process.

@@

We can now put these processes together to type the courier system.
Specifying the types of the channels is not required: Agda can infer them from the types of the processes.
